---
title: "Vital status value inconsistencies"
author: "Haley Hunter-Zinck"
date: "`r Sys.setenv(TZ='America/Los_Angeles'); format(Sys.time(), '%B %d, %Y at %I:%M %p %Z')`"
output: html_document
parameters:
  synid_file_patient: x
  synid_version_patient: x
---

## Overview

We look for discrepancies in vital status information submitted for main GENIE consortium releases.  

We check the following variables:

* YEAR_DEATH
* INT_DOD
* DEAD

for inconsistencies with any of the following rules: 

1. If the YEAR_DEATH is redacted ('>89' or '<18') then INT_DOD must also be redacted ('<6570' or '>32485') and vice versa.
2. If YEAR_DEATH or INT_DOD is an unredacted text value ('Unknown', 'Not Collected', 'Not Applicable', 'Not Released'), INT_DOD and YEAR_DEATH must hold the same text value.  
3. If YEAR_DEATH or INT_DOD is 'Not Applicable', DEAD must be False.
4. If INT_DOD is a redacted or unredacted number, DEAD must be True.
5. YEAR_DEATH AND INT_DOD must not be NA.
6. DEAD must be 'False' or 'True'

## Parameters

Synapse ID of patient file: `r synid_file_patient`

Synapse version of patient file: `r synid_version_patient`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

library(synapser)
invisible(synLogin())
library(glue)
library(dplyr)

```

```{r functions, include = F}
#' Download and load data stored in csv or other delimited format on Synapse
#' into an R data frame.
#' 
#' @param synapse_id Synapse ID
#' @version Version of the Synapse entity to download.  NA will load current
#' version
#' @param set Delimiter for file
#' @param na.strings Vector of strings to be read in as NA values
#' @param header TRUE if the file contains a header row; FALSE otherwise.
#' @param check_names TRUE if column names should be modified for compatibility 
#' with R upon reading; FALSE otherwise.
#' @return data frame
get_synapse_entity_data_in_csv <- function(synapse_id, 
                                           version = NA,
                                           sep = ",", 
                                           na.strings = c("NA"), 
                                           header = T,
                                           check_names = F, 
                                           comment.char = "#") {
  
  if (is.na(version)) {
    entity <- synGet(synapse_id)
  } else {
    entity <- synGet(synapse_id, version = version)
  }
  
  data <- read.csv(entity$path, stringsAsFactors = F, 
                   na.strings = na.strings, sep = sep, check.names = check_names,
                   header = header, comment.char = comment.char)
  return(data)
}
```

```{r read, include = F}

data <- get_synapse_entity_data_in_csv(synid_file_patient, version = synid_version_patient, sep = "\t", na.strings = c("NA", ""))

```


## Vital data summary

Number of patients = `r nrow(data)` 

Unique non-numeric values for YEAR_DEATH:
```{r YEAR_DEATH values, include = T}
data %>% 
  filter(!(grepl(pattern = "[0-9]{4}", x = YEAR_DEATH))) %>% 
  select(YEAR_DEATH) %>% 
  group_by(YEAR_DEATH) %>% 
  count() %>%
  arrange(desc(n))
```

Unique non-numeric values for INT_DOD:
```{r INT_DOD values, include = T}
data %>% 
  filter(!(grepl(pattern = "[0-9]{4}", x = INT_DOD))) %>% 
  select(INT_DOD) %>% 
  group_by(INT_DOD) %>% 
  count() %>%
  arrange(desc(n))
```

Unique values for DEAD:
```{r DEAD values, include = T}
data %>% 
  select(DEAD) %>% 
  group_by(DEAD) %>% 
  count() %>%
  arrange(desc(n))
```

## Checks for vital information inconsistencies

### Check 1

```{r redaction, include = F}
data_err_red <- data %>%
  filter(grepl(pattern = "[<>]", x = YEAR_DEATH) & !grepl(pattern = "[<>]", x = INT_DOD) |
           !grepl(pattern = "[<>]", x = YEAR_DEATH) & grepl(pattern = "[<>]", x = INT_DOD)) %>%
  select(YEAR_DEATH, INT_DOD, CENTER)
```

Number of inconsistent redactions: `r nrow(data_err_red)`

Centers with inconsistent redactions:
```{r redaction centers, include = T}
data_err_red %>% select(CENTER) %>% distinct()
```

Sample of inconsistent redaction entries: 
```{r redaction examples, include = T}
head(data_err_red %>% select(YEAR_DEATH, INT_DOD) %>% distinct())
```

### Check 2

```{r text value match, include = F}
strs <- c("Unknown", "Not Collected", "Not Applicable", "Not Released")
data_err_text <- data %>%
  filter((is.element(YEAR_DEATH, strs) | is.element(INT_DOD, strs)) & YEAR_DEATH != INT_DOD) %>%
  select(YEAR_DEATH, INT_DOD, CENTER)
```

Number of inconsistent text values: `r nrow(data_err_text)`

Centers with inconsistent text values:
```{r text centers, include = T}
data_err_text %>% select(CENTER) %>% distinct()
```

Sample of inconsistent unredacted text values:
```{r text examples, include = T}
head(data_err_text %>% select(YEAR_DEATH, INT_DOD) %>% distinct())
```

### Check 3

```{r NA DEAD F, include = F}
strs_true <- c("TRUE", "True")
data_err_dead_na <- data %>% 
  filter((INT_DOD == 'Not Applicable' | YEAR_DEATH == 'Not Applicable') & is.element(DEAD, strs_true)) %>%
  select(INT_DOD, YEAR_DEATH, DEAD, CENTER)
```

Number of inconsistent DEAD values for 'Not Applicable' vital info: `r nrow(data_err_dead_na)`

Centers with inconsistent DEAD values for 'Not Applicable' vital info:
```{r NA dead centers, include = T}
data_err_dead_na %>% select(CENTER) %>% distinct()
```

Distinct patterns of inconsistent DEAD values with 'Not Applicable' vitals:
```{r dead NA examples, include = T}
data_err_dead_na %>% select(INT_DOD, YEAR_DEATH, DEAD) %>% distinct()
```

### Check 4

```{r red DEAD T, include = F}
strs_false <- c("FALSE", "False")
data_err_dead_red <- data %>% 
  filter((grepl(pattern = "[<>]", x = YEAR_DEATH) | grepl(pattern = "[<>]", x = INT_DOD)) & is.element(DEAD, strs_false)) %>%
  select(INT_DOD, YEAR_DEATH, DEAD, CENTER)
```

Number of inconsistent DEAD values for redacted vitals: `r nrow(data_err_dead_red)`

Centers with inconsistent DEAD values for redacted vitals:
```{r redacted dead centers, include = T}
data_err_dead_red %>% select(CENTER) %>% distinct()
```

Distinct patterns of inconsistent DEAD values with redacted vitals:
```{r dead red examples, include = T}
data_err_dead_red %>% select(INT_DOD, YEAR_DEATH, DEAD) %>% distinct()
```

